---
title: "R lecture 3: Meta-analysis in R I"
subtitle: "SKI3011 · escalc(), rma(), forest plots, proportions and correlations"
# This page is also the lecture script: scripts/render-r-lectures.sh turns it into an .R file
# in which all text below becomes comments, so students can run it line by line in RStudio.
execute:
  echo: true
  warning: false
  message: false
---

**How to use this lecture.** Download the R script, open it in RStudio and run it line by line (Cmd+Enter on a Mac, Ctrl+Enter on Windows). All explanations on this page are in the script as comments. Check that you get the same output as shown here. Quiz 2 asks you to read and interpret exactly this kind of code and output.

[Download the R script](../materials/ski3011/week-3/r-lecture-3.R){.btn .btn-primary}

## Packages

`install.packages()` installs a package once on your computer; `library()` loads it in every new R session.

```{r}
# install.packages(c("metafor", "metadat"))   # run once, remove the # to run it
library(metafor)   # meta-analysis functions: escalc(), rma(), forest()
library(metadat)   # example datasets, such as dat.bcg
```

## The BCG data

13 trials of the BCG vaccine against tuberculosis. `tpos` and `tneg` are the TB-positive and TB-negative people in the vaccinated group, `cpos` and `cneg` the same in the control group: the four cells of the 2×2 table. `ablat` is the absolute latitude of the trial site; we use it in week 6.

```{r}
head(dat.bcg[, c("author", "year", "tpos", "tneg", "cpos", "cneg", "ablat")], 4)
```

## Step 1: one effect size per study

`escalc()` adds two columns to the data: `yi`, the **log** risk ratio, and `vi`, its sampling variance (SE²). Ratios are always stored on the log scale.

```{r}
dat <- escalc(measure = "RR",                              # effect size: risk ratio (also "OR", "RD", "MD", "SMD", "PR", "ZCOR")
              ai = tpos, bi = tneg, ci = cpos, di = cneg,  # the four cells of the 2x2 table
              data = dat.bcg,
              slab = paste(author, year))                  # study labels for the output and the forest plot
head(dat[, c("author", "year", "yi", "vi")], 4)
```

## Step 2: random-effects meta-analysis

`rma()` fits a **random-effects** model by default, with τ² estimated by REML. `tau^2`, `I^2` and the Q test describe heterogeneity (week 5). `estimate` is the pooled **log** risk ratio: −0.71.

```{r}
res <- rma(yi, vi, data = dat)   # random-effects model (default method: REML)
print(res, digits = 2)
```

## Step 3: back to the risk ratio scale

`pred` = exp(−0.71) = 0.49: vaccinated people have about half the risk of TB (95% CI 0.34 to 0.70). `pi.lb` and `pi.ub` are the 95% prediction interval: the range in which the true effect of a new trial is expected to fall (week 5).

```{r}
predict(res, transf = exp, digits = 2)   # exp() turns the log risk ratio back into a risk ratio
```

## Step 4: forest plot

`atransf = exp` shows risk ratios on the axis instead of log risk ratios. Box size is the weight of each trial; the diamond is the pooled estimate with its 95% CI.

```{r}
#| fig-height: 6
#| fig-width: 9
forest(res, atransf = exp, header = TRUE)
```

## Fixed effect versus random effects

The fixed-effect model assumes one common true effect. Its CI is much narrower (0.60 to 0.70) because it ignores the large between-trial variation: with heterogeneity this big, the random-effects model is the honest choice.

```{r}
res_fe <- rma(yi, vi, data = dat, method = "FE")   # fixed-effect model
predict(res_fe, transf = exp, digits = 2)
```

## Proportions

35 studies on the proportion of COVID-19 patients with loss of smell (dat.hannum2020). About half of the patients (0.49, 95% CI 0.41 to 0.57) lose their sense of smell. Note the enormous heterogeneity: the studies used objective and subjective measurements (we split them in week 6). For proportions near 0 or 1 a transformation such as `"PLO"` (logit) is better.

```{r}
dat_pr <- escalc(measure = "PR", xi = xi, ni = ni, data = dat.hannum2020)   # xi = events, ni = sample size
res_pr <- rma(yi, vi, data = dat_pr)
print(res_pr, digits = 2)
```

## Correlations

16 studies on the correlation between conscientiousness and medication adherence (dat.molloy2014). Correlations are pooled after transformation to Fisher's z and transformed back with `transf.ztor`. More conscientious patients adhere slightly better to their medication: r = 0.15 (95% CI 0.09 to 0.21). The prediction interval (−0.04 to 0.32) shows that in some settings there may be no association at all.

```{r}
dat_r <- escalc(measure = "ZCOR", ri = ri, ni = ni, data = dat.molloy2014)   # ri = correlation, ni = sample size
res_r <- rma(yi, vi, data = dat_r)
predict(res_r, transf = transf.ztor, digits = 2)   # back from Fisher's z to a correlation
```

## Check yourself

Go to the [concept list and practice quiz of week 3](week-3.qmd) and try the practice quiz without looking at this page.
