---
title: "R lecture 3: Meta-analysis in R I"
subtitle: "SKI3011 · escalc(), rma(), forest plots, proportions and correlations"
# One source, two outputs: this web page and a slide deck (revealjs).
# Text inside ::: {.content-visible unless-format="revealjs"} only appears on the web page.
format:
  html:
    code-annotations: below
    code-tools:
      source: false
  revealjs:
    output-file: r-lecture-3-slides.html
    slide-number: true
    code-annotations: below
    scrollable: true
    smaller: true
    theme: default
execute:
  echo: true
  warning: false
  message: false
---

::: {.content-visible unless-format="revealjs"}
**How to use this page.** This is the full lecture: explanation, R code and the real output. In the carbon-based weeks you type and run every line yourself in RStudio: download the script below, run it line by line, and check that you get the same output. Quiz 2 asks you to read and interpret exactly this kind of code and output.

* [Download the R script](../materials/ski3011/week-3/r-lecture-3.R)
* [Open as slides](r-lecture-3-slides.html)
:::

## Packages

```{r}
# install.packages(c("metafor", "metadat"))   # once on your computer
library(metafor)   # meta-analysis functions: escalc(), rma(), forest()
library(metadat)   # example datasets, such as dat.bcg
```

::: {.content-visible unless-format="revealjs"}
`install.packages()` installs a package once; `library()` loads it in every new R session.
:::

## The BCG data

13 trials of the BCG vaccine against tuberculosis.

```{r}
head(dat.bcg[, c("author", "year", "tpos", "tneg", "cpos", "cneg", "ablat")], 4)
```

::: {.content-visible unless-format="revealjs"}
`tpos` and `tneg` are the TB-positive and TB-negative people in the vaccinated group, `cpos` and `cneg` the same in the control group: the four cells of the 2×2 table. `ablat` is the absolute latitude of the trial site; we use it in week 6.
:::

## Step 1: one effect size per study

```{r}
dat <- escalc(measure = "RR",                              # <1>
              ai = tpos, bi = tneg, ci = cpos, di = cneg,  # <2>
              data = dat.bcg,
              slab = paste(author, year))                  # <3>
head(dat[, c("author", "year", "yi", "vi")], 4)
```
1. The effect size: the risk ratio. Other options: `"OR"`, `"RD"`, `"MD"`, `"SMD"`, `"PR"`, `"ZCOR"`.
2. The four cells of the 2×2 table.
3. Study labels for the output and the forest plot.

::: {.content-visible unless-format="revealjs"}
`escalc()` adds two columns: `yi`, the **log** risk ratio, and `vi`, its sampling variance (SE²). Ratios are always stored on the log scale.
:::

## Step 2: random-effects meta-analysis

```{r}
res <- rma(yi, vi, data = dat)
print(res, digits = 2)
```

::: {.content-visible unless-format="revealjs"}
* `rma()` fits a **random-effects** model by default, with τ² estimated by REML.
* `tau^2`, `I^2` and the Q test describe heterogeneity (week 5).
* `estimate` is the pooled **log** risk ratio: −0.71.
:::

## Step 3: back to the risk ratio scale

```{r}
predict(res, transf = exp, digits = 2)
```

::: {.content-visible unless-format="revealjs"}
`pred` = exp(−0.71) = 0.49: vaccinated people have about half the risk of TB (95% CI 0.34 to 0.70). `pi.lb` and `pi.ub` are the 95% prediction interval: the range in which the true effect of a new trial is expected to fall (week 5).
:::

## Step 4: forest plot

```{r}
#| fig-height: 6
#| fig-width: 9
forest(res, atransf = exp, header = TRUE)
```

::: {.content-visible unless-format="revealjs"}
`atransf = exp` shows risk ratios on the axis instead of log risk ratios. Box size is the weight of each trial; the diamond is the pooled estimate with its 95% CI.
:::

## Fixed effect versus random effects

```{r}
res_fe <- rma(yi, vi, data = dat, method = "FE")
predict(res_fe, transf = exp, digits = 2)
```

::: {.content-visible unless-format="revealjs"}
The fixed-effect model assumes one common true effect. Its CI is much narrower (0.60 to 0.70) because it ignores the large between-trial variation: with heterogeneity this big, the random-effects model is the honest choice.
:::

## Proportions

35 studies on the proportion of COVID-19 patients with loss of smell (dat.hannum2020).

```{r}
dat_pr <- escalc(measure = "PR", xi = xi, ni = ni, data = dat.hannum2020)   # <1>
res_pr <- rma(yi, vi, data = dat_pr)
print(res_pr, digits = 2)
```
1. `xi` = number of events, `ni` = sample size; `"PR"` pools the raw proportions.

::: {.content-visible unless-format="revealjs"}
About half of the patients (0.49, 95% CI 0.41 to 0.57) lose their sense of smell. Note the enormous heterogeneity: the studies used objective and subjective measurements (we split them in week 6). For proportions near 0 or 1 a transformation such as `"PLO"` (logit) is better.
:::

## Correlations

16 studies on the correlation between conscientiousness and medication adherence (dat.molloy2014).

```{r}
dat_r <- escalc(measure = "ZCOR", ri = ri, ni = ni, data = dat.molloy2014)   # <1>
res_r <- rma(yi, vi, data = dat_r)
predict(res_r, transf = transf.ztor, digits = 2)                               # <2>
```
1. Correlations are transformed to Fisher's z before pooling.
2. `transf.ztor` transforms the pooled z back to a correlation.

::: {.content-visible unless-format="revealjs"}
More conscientious patients adhere slightly better to their medication: r = 0.15 (95% CI 0.09 to 0.21). The prediction interval (−0.04 to 0.32) shows that in some settings there may be no association at all.
:::

::: {.content-visible unless-format="revealjs"}
## Check yourself

Go to the [concept list and practice quiz of week 3](week-3.qmd) and try the practice quiz without looking at this page.
:::
