---
title: "R lecture 2: R basics"
subtitle: "SKI3011 · objects, vectors, data frames, reading data, data types, plots"
# This page is also the lecture script: scripts/render-r-lectures.sh turns it into an .R file
# in which all text below becomes comments, so students can run it line by line in RStudio.
execute:
  echo: true
  warning: false
  message: false
---

**How to use this lecture.** Download the R script, open it in RStudio and run it line by line (Cmd+Enter on a Mac, Ctrl+Enter on Windows). All explanations on this page are in the script as comments. Check that you get the same output as shown here. Quiz 1 asks you to read and interpret exactly this kind of code and output.

[Download the R script](../materials/ski3011/week-2/r-lecture-2.R){.btn .btn-primary} [Download the practice file bcg.csv](../materials/ski3011/week-2/bcg.csv){.btn .btn-outline-primary}

```{r}
#| include: false
# the web page reads bcg.csv from the materials folder; students keep it in their own folder
knitr::opts_knit$set(root.dir = "../materials/ski3011/week-2")
```

## Why R?

R is a program for statistical computing, not just for meta-analyses. To learn to do a meta-analysis in R, we first need the basics. A general principle of responsible research: keep only two files, the raw data file and the script. Never change the raw data by hand without logging it: every change is a line in the script.

## R as a calculator

```{r}
4 + 3
4 * 2
4^2
log(4)    # natural logarithm
exp(4)    # the inverse of log()
sqrt(4)
```

## Vectors and functions

A vector combines values that belong together, like a column in a spreadsheet: the ages of participants, or the effect sizes of the studies in a meta-analysis. `c()` stands for combine. A function such as `mean()` takes arguments between brackets. Use `?` to open the help page of a function.

```{r}
c(9, 3, 6)
mean(c(9, 3, 6))
sd(c(9, 3, 6))
sort(c(9, 3, 6), decreasing = TRUE)   # an extra argument changes what the function does
# ?sort                               # remove the # to open the help page
```

## Objects

A vector, a dataset or a single value can be stored in an object with `<-`. You choose the name yourself. The object appears in the Environment window (top right in RStudio).

```{r}
a <- c(9, 3, 6)
a
b <- 2 + 2
a / b      # every value of a is divided by b
z <- a / b
z
```

## Data frames

A data frame is a table: one row per study, one column per variable. Here we make the meta-analysis dataset you know from SKI3010 by hand: 11 studies on environmental tobacco smoke at home and bladder cancer, with the log odds ratio and its standard error.

```{r}
lnOR <- c(-0.223143551, -0.198450939, -0.162518929, -0.162518929, -0.116533816, 0.009950331,
          0.086177696, 0.139761942, 0.165514438, 0.336472237, 0.457424847)
se <- c(0.38457745, 0.301271952, 0.200831965, 0.279480271, 0.359347184, 0.428673775,
        0.247347883, 0.162072911, 0.251023247, 0.3350916, 0.293754916)
firstauthor <- c("Alberg (2)", "Bjerregaard", "Jiang", "Burch", "NLCS", "Kabat",
                 "Baris", "Zheng", "Samanic", "Alberg (1)", "Tao")
year <- c(2007, 2006, 2007, 1989, 2002, 1986, 2009, 2012, 2006, 2007, 2010)
dat <- data.frame(firstauthor, year, lnOR, se)
dat
names(dat)                # the column names
dat$lnOR                  # $ selects one column
dat$year[2]               # [ ] selects an element: the second year
dat[dat$year > 2006, ]    # the rows of studies published after 2006
round(exp(dat$lnOR), 2)   # back from log odds ratios to odds ratios
# View(dat)               # opens the data in a spreadsheet view
```

## Reading a csv file

Most of your data will come from a spreadsheet saved as csv. Download the practice file bcg.csv and put it in the folder where you keep your script. `getwd()` shows the folder R works in; in RStudio you can change it via Session, Set Working Directory, To Source File Location.

European csv files use a semicolon as separator and a comma as decimal sign. If you forget that, R reads everything into one column:

```{r}
# getwd()                    # which folder is R working in?
wrong <- read.csv("bcg.csv")
head(wrong, 2)               # one column, everything glued together
```

Tell R which separator and decimal sign the file uses:

```{r}
bcg <- read.csv("bcg.csv", sep = ";", dec = ",")
head(bcg, 3)
dim(bcg)        # number of rows (studies) and columns (variables)
```

## Data types

R distinguishes numeric values (for calculations), logical values (TRUE or FALSE), character values (text) and factors (categories). `class()` tells you which type an object is.

```{r}
a <- 44 / 5
class(a)
b <- 7 > 4
b
class(b)
class(bcg$author)
pets <- c("cat", "dog", "cat", "bird", "cat")
class(pets)
pets <- as.factor(pets)   # a factor: a categorical variable with levels
levels(pets)
```

## Some simple statistics

R has built-in datasets. `airquality` contains daily air measurements in New York. Does ozone concentration depend on temperature? In `lm()` the tilde `~` separates the outcome from the explanatory variable.

```{r}
dim(airquality)
head(airquality, 3)
fit <- lm(Ozone ~ Temp, data = airquality)
summary(fit)
```

## Plots

The built-in dataset `iris` has flower measurements of three species. Arguments such as `main`, `xlab`, `pch` and `col` change the title, axis labels, symbols and colours.

```{r}
hist(iris$Petal.Length, main = "Histogram of petal length", xlab = "Petal length")
```

```{r}
boxplot(Petal.Length ~ Species, data = iris,
        main = "Petal length by species", xlab = "Species", ylab = "Petal length")
```

```{r}
plot(iris$Petal.Length, iris$Sepal.Length, pch = 20, col = "darkorange",
     main = "Petal length and sepal length", xlab = "Petal length", ylab = "Sepal length")
abline(lm(Sepal.Length ~ Petal.Length, data = iris))   # add the regression line
```

## Saving your data

`write.csv()` saves a data frame as a csv file in your working directory; `read.csv()` reads it back. This line is not run on the website.

```{r}
#| eval: false
write.csv(bcg, "my_bcg_data.csv", row.names = FALSE)
my_data <- read.csv("my_bcg_data.csv")
```

## Now your own data

Import your own extraction sheet (saved as csv) with `read.csv()` and check it with `head()`, `dim()` and `class()`.

## Check yourself

Go to the [concept list and practice quiz of week 2](week-2.qmd) and try the practice quiz without looking at this page.
