---
title: "R lecture 5: Heterogeneity and publication bias"
subtitle: "SKI3011 · tau², I², prediction intervals, influence, cumulative meta-analysis, funnel plots"
# This page is also the lecture script: scripts/render-r-lectures.sh turns it into an .R file
# in which all text below becomes comments, so students can run it line by line in RStudio.
execute:
  echo: true
  warning: false
  message: false
---

**How to use this lecture.** Download the R script, open it in RStudio and run it line by line (Cmd+Enter on a Mac, Ctrl+Enter on Windows). All explanations on this page are in the script as comments. Check that you get the same output as shown here. Quiz 4 asks you to read and interpret exactly this kind of code and output.

[Download the R script](../materials/ski3011/week-5/r-lecture-5.R){.btn .btn-primary}

## Packages and data

We start again with the BCG trials of lecture 3.

```{r}
library(metafor)
library(metadat)
dat <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg,
              data = dat.bcg, slab = paste(author, year))
res <- rma(yi, vi, data = dat)
```

## How much heterogeneity?

* `tau^2` (τ²): the variance of the true effects between studies; `tau` is its square root.
* `I^2`: the share of the total variability due to real differences between studies rather than chance.
* `H^2`: total variability divided by sampling variability (1 means no heterogeneity).
* The Q test: is there more variation than chance would explain? It has little power when there are few studies.

```{r}
print(res, digits = 2)
confint(res, digits = 2)   # confidence intervals for tau^2 and I^2
```

I² is 92% (95% CI 82% to 98%): the effect of BCG differs strongly between trials.

## The prediction interval

The confidence interval tells you how precisely the average effect is estimated. The prediction interval (`pi.lb`, `pi.ub`) tells you where the true effect of a new trial is expected to fall. With large τ² it is much wider.

```{r}
predict(res, transf = exp, digits = 2)
```

On average BCG halves the risk (RR 0.49), but in a new setting the effect could be anything from strong protection (0.15) to no protection or even harm (1.55).

## Is one study driving the result?

`leave1out()` repeats the meta-analysis k times, each time without one study. If the pooled estimate or I² changes a lot when one study is left out, that study is influential.

```{r}
leave1out(res, transf = exp, digits = 2)
```

`influence()` gives formal diagnostics: studies marked with an asterisk are influential.

```{r}
#| fig-height: 7
inf <- influence(res)
print(inf, digits = 2)
plot(inf)
```

## Cumulative meta-analysis

A cumulative meta-analysis adds the studies one by one, here in order of publication year. It shows how the evidence developed over time.

```{r}
#| fig-height: 5
#| fig-width: 9
cum <- cumul(res, order = dat$year)
forest(cum, atransf = exp, header = TRUE)
```

## Publication bias: the magnesium story

16 trials of intravenous magnesium after a heart attack, outcome death (dat.egger2001). The small early trials suggested a large benefit. The last trial, ISIS-4 (1995), included more than 58,000 patients and found no benefit.

```{r}
mag <- escalc(measure = "OR", ai = ai, n1i = n1i, ci = ci, n2i = n2i,
              data = dat.egger2001, slab = paste(study, year))
res_mag <- rma(yi, vi, data = mag)
predict(res_mag, transf = exp, digits = 2)
```

## The funnel plot

A funnel plot shows each study's effect against its standard error. Without bias, small studies (at the bottom) scatter symmetrically around the pooled estimate. Here the small studies all sit on the side of a large benefit: the funnel is asymmetric.

```{r}
#| fig-height: 5
funnel(res_mag, atransf = exp, xlab = "Odds ratio (log scale)")
```

## Egger's test

`regtest()` tests funnel plot asymmetry. The limit estimate is the expected effect of an infinitely large study (standard error → 0).

```{r}
regtest(res_mag)
```

The test is clearly significant, and the limit estimate is close to 0 on the log scale (an OR of about 1): exactly what ISIS-4 found.

## Trim-and-fill

`trimfill()` estimates how many studies are "missing" and what the pooled estimate would be if they were added. The filled studies appear as open circles in the funnel plot.

```{r}
#| fig-height: 5
tf <- trimfill(res_mag)
predict(tf, transf = exp, digits = 2)
funnel(tf, atransf = exp, xlab = "Odds ratio (log scale)")
```

After filling 7 studies the OR rises from 0.46 to 0.68 and is no longer statistically significant. Trim-and-fill is a sensitivity analysis, not a correction: it shows how fragile the result is.

## Small-study effects are not always publication bias

Funnel plot asymmetry means that small studies show different effects than large ones. Publication bias is one explanation; others are lower quality of small trials, different patients, or chance. With fewer than about 10 studies these tests have little power.

## Check yourself

Go to the [concept list and practice quiz of week 5](week-5.qmd) and try the practice quiz without looking at this page.
